---
title: "Data wrangling"
output: pdf_document
---

## Load Packages
```{r warning=FALSE, message = FALSE}
require(mosaic)
require(tidyverse)
require(lubridate)
require(rvest)
require(lme4)
```

## Load Datasets
```{r message=FALSE, warning=FALSE}
wd <- getwd()
regions <- read_csv(paste0(wd,"/Data/NOC_Region.csv"))
athletes <- read_csv(paste0(wd,"/Data/Olympics_Athletes.csv"))
countrycode <- read_csv(paste0(wd,"/Data/ISOCountryCode.csv"))
temphistory <- readxl::read_xlsx(paste0(wd,"/Data/Historical_Temp_Data.xlsx"), sheet = 2)
olymweat <- readxl::read_xlsx(paste0(wd,"/Data/Olympics_Weather_Data.xlsx"))
inout <- Class.csv <- read_csv(paste0(wd, "/Data/indooroutdoorClass.csv"))
```

## Data Previews
```{r}
head(regions)
head(athletes)
head(countrycode)
head(temphistory)
head(olymweat, 20)
head(inout)
```

## Combine countrycode and temphistory data
```{r}
# combine countrycode and temphistory
countrytemp <- left_join(temphistory, countrycode, by = c("ISO_3DIGIT" = "alpha-3")) %>%
  select(-c(16:24), -`ISO_3DIGIT`)

# change name of each month column so that wrangling is easier later on
names(countrytemp)[c(1:12)] <- c(1:12)
```

## Clean up olymweat dates 
```{r}
# clean up dates column
olymweat2 <- olymweat %>%
  rename("StartDate" = `Start Date`,
         "EndDate" = `End Date`) %>%
  mutate(StartDate = as.character(StartDate),
         EndDate = as.character(EndDate),
         StartMonth = strsplit(StartDate, split = "-")[[1]][2],
         EndMonth = strsplit(EndDate, split = "-")[[1]][2],
         StartDay = strsplit(StartDate, split = "-")[[1]][3],
         EndDay = strsplit(EndDate, split = "-")[[1]][3]) %>%
  select(-StartDate, -EndDate) 

head(olymweat2)
```

## Combine athletes and region
```{r}
athletes2 <- athletes %>%
  left_join(regions, by = c('NOC' = 'NOC')) %>%
  select(-notes, -NOC, -Games)
```

## Wrangling elevation data 
```{r}
wikiTable <- function(source) {
  read_html(source)%>%
    html_nodes("table.wikitable") %>%
    html_table(fill=T)%>%
    magrittr::extract2(1)
}
# Read in wikipedia table for average elevation per country
elevation <- wikiTable("https://en.wikipedia.org/wiki/List_of_countries_by_average_elevation#cite_note-1")

# Clean up elevation data
elelist <- strsplit(elevation$Elevation, "m")
vec <- c()
for (i in 1:length(elelist)){
  vec[i] <- elelist[[i]][1]
}

# Clean up dataframe
elevation$Elevation <- vec
elevation <- elevation %>%
  mutate(Elevation = parse_number(Elevation))
```



## Combine olymweat2 and athletes and filter out indoor sports
```{r}
olympics1 <- left_join(athletes2, olymweat2, by = c("Year" = "Year", "Season" = "Season")) %>%
  select(-City.y) %>%
  rename(City = City.x) %>%
  filter(!(Year %in% c(1896,1900,1904,1906,1908,1912,1920,1924,1928) & Season == "Summer")) %>%
  mutate(CityTemp1 = `Hist Avg Temp Mon1 [C]` + `Diff Avg Temp Mon1 [C]`,
         CityTemp2 = `Hist Avg Temp Mon2 [C]` + `Diff Avg Temp Mon2 [C]`) %>%
  select(-`Hist Avg Temp Mon1 [C]`, -`Hist Avg Temp Mon2 [C]`, 
         -`Diff Avg Temp Mon1 [C]`, -`Diff Avg Temp Mon2 [C]`) %>%
  left_join(inout, by = c("Sport" = "Sport")) %>%
  # filtering out indoor sports
  filter(isindoor == 0)
```


## Combine olympics1 with countrytemp
```{r}
olympics2 <- left_join(olympics1, countrytemp, by = c("region" = "name"))

# create vector of the starting and ending months of each olympic
startmon <- as.character(as.integer(olympics1$StartMonth))
endmon <- as.character(as.integer(olympics1$EndMonth))

# loop through olympics2 to gather the temperature of months that corresponds to the olympic months for each athlete's home country
# store this data in histtemp1 and histtemp2

histtemp1 <- c()
histtemp2 <- c()
for (i in 1:nrow(olympics2)){
  histtemp1[i] <- olympics2[[i,startmon[i]]]
  histtemp2[i] <- olympics2[[i,endmon[i]]]
}

# add create
olympics2$histtemp1 <- histtemp1
olympics2$histtemp2 <- histtemp2

# get rid of temperature data of every month
olympics2 <- olympics2 %>%
  select(-`1`,-`2`,-`3`,-`4`,-`5`,-`6`,-`7`,-`8`,-`9`,-`10`,-`11`,-`12`, -Annual_temp)

olympics2
```


## Calculate difference between host city temperature and home country temperature
```{r}
# calculate the ratio of days in the first and second month
olympics3 <- olympics2 %>%
  mutate(NumDayMon1 = ifelse(as.integer(EndDay) > `Number of Days`, `Number of Days`, `Number of Days` - as.integer(EndDay)),
         NumDayMon2 = `Number of Days` - NumDayMon1,
         RatioMon1 = NumDayMon1/`Number of Days`,
         RatioMon2 = 1 - RatioMon1,
         CityTemp2 = ifelse(is.na(CityTemp2), 0, CityTemp2)) %>%
# calculate the total difference
  mutate(tempdiff = (CityTemp1 - histtemp1)*RatioMon1 + (CityTemp2 - histtemp2)*RatioMon2) %>%
  
# get rid of columns that were used in the calculation for temperature difference
  select(-CityTemp1, -CityTemp2, -histtemp1, -histtemp2, -NumDayMon1, -NumDayMon2, -RatioMon1, -RatioMon2)
```

## Combine elevation data with olympics3 and compute elevation difference
```{r}
olympics4 <- olympics3 %>%
  left_join(elevation, by = c("region" = "Country")) %>%
  mutate(elevdiff = Altitude - Elevation) %>% 
  select(-Elevation, -Altitude)
```

## Clean up medal data
```{r}
olympics5 <- olympics4 %>%
  mutate(Medal = ifelse(is.na(Medal), "0", Medal),
         Medal = ifelse(Medal == "Bronze", "1", Medal),
         Medal = ifelse(Medal == "Silver", "2", Medal),
         Medal = ifelse(Medal == "Gold", "3", Medal),
         Medal = as.integer(Medal)) 
  
  
totalmedals <- olympics5 %>%
  group_by(Year,Season) %>%
  summarise(TotMed = sum(Medal))
totalmedals
```

## Calculate percentage of medals won by each country
```{r}
olympics6 <- olympics5 %>%
  group_by(Year, Season, region) %>%
  summarise(medalswon = sum(Medal)) %>%
  left_join(totalmedals, by = c("Year" = "Year", "Season" = "Season")) %>%
  mutate(medratio = medalswon / TotMed)
```

## Grouped Countries by getting rid of distinctions between each athletes
```{r}
olympics7 <- olympics4 %>% 
  select(Year, Season, region, tempdiff, elevdiff)

duplicated <- olympics4 %>% 
  select(Year, Season, region, tempdiff, elevdiff) %>%
  duplicated() 

olympics8 <- olympics7[!duplicated,]
```

## Combined temperature/elevation data with medal data
```{r}
olympics9 <- olympics6 %>%
  left_join(olympics8, by = c("Year" = "Year", "Season" = "Season", "region" = "region"))
```


## save olympics9 as R.data
```{r}
save(olympics9, "olympics9", file = "olympics9.Rdata")
```

















