---
title: "Data wrangling"
output: pdf_document
---

## Document Options
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE)
```

## Load Packages
```{r}
require(mosaic)
require(tidyverse)
require(lubridate)
require(rvest)
require(lme4)
```

## Load Datasets
```{r message=FALSE, warning=FALSE}
regions <- read_csv("Data/NOC_Region.csv")
athletes <- read_csv("Data/Olympics_Athletes.csv")
countrycode <- read_csv("Data/ISOCountryCode.csv")
temphistory <- readxl::read_xlsx("Data/Historical_Temp_Data.xlsx", sheet = 2)
olymweat <- readxl::read_xlsx("Data/Olympics_Weather_Data.xlsx")
inout <- Class.csv <- read_csv("Data/indooroutdoorClass.csv")
```


## Wrangling elevation data 
```{r}
# Create function to extract elevation data from wikipedia
wikiTable <- function(source) {
  read_html(source)%>%
    html_nodes("table.wikitable") %>%
    html_table(fill=T)%>%
    magrittr::extract2(1)
}

# Read in wikipedia table for average elevation per country
elevation <- wikiTable("https://en.wikipedia.org/wiki/List_of_countries_by_average_elevation#cite_note-1")

# Clean up format of elevation value
elelist <- strsplit(elevation$Elevation, "m")
vec <- c()
for (i in 1:length(elelist)){
  vec[i] <- elelist[[i]][1]
}
elevation$Elevation <- vec
elevation <- elevation %>%
  mutate(Elevation = parse_number(Elevation))

# Change format of name of countries
pos <- which(elevation$Country %in% c("Trinidad and Tobago","United Kingdom","United States"))
elevation[pos,1] <- c("Trinidad","UK", "USA")
```

## Glimpse Data 
```{r}
glimpse(regions)
glimpse(athletes)
glimpse(countrycode)
glimpse(temphistory)
glimpse(olymweat)
glimpse(inout)
glimpse(elevation)
```

## Combine countrycode and temphistory data
```{r}
# Join countrycode and temphistory 
countrytemp <- left_join(temphistory, countrycode, by = c("ISO_3DIGIT" = "alpha-3")) %>%
  select(-c(16:24), -`ISO_3DIGIT`) %>%

# Rename each month column so that wrangling is easier later on
  plyr::rename(c("Jan_Temp" = "1", "Feb_temp" = "2",
                 "Mar_temp" = "3", "Apr_Temp" = "4",
                 "May_temp" = "5", "Jun_Temp" = "6",
                 "July_Temp" = "7", "Aug_Temp" = "8",
                 "Sept_temp" = "9", "Oct_temp" = "10",
                 "Nov_Temp" = "11", "Dec_temp" = "12"))
```

## Clean up countrytemp 
```{r}
# Rename some countries to standardize country name format
countrytemp <- countrytemp %>%
  mutate(name = ifelse(name == "Bolivia (Plurinational State of)", "Bolivia", name),
         name = ifelse(name == "Czechia", "Czech Republic", name),
         name = ifelse(name == "Korea (Republic of)", "South Korea", name),
         name = ifelse(name == "Russian Federation", "Russia", name),
         name = ifelse(name == "Trinidad and Tobago", "Trinidad", name),
         name = ifelse(name == "United Kingdom of Great Britain and Northern Ireland", "UK", name),
         name = ifelse(name == "United States of America", "USA", name),
         name = ifelse(name == "Venezuela (Bolivarian Republic of)", "Venezuela", name),
         name = ifelse(name == "Viet Nam", "Vietnam", name))
```

## Clean up dates of olymweat 
```{r}
# Create new variables for starting/ending months and days
olymweat2 <- olymweat %>%
  rename("StartDate" = `Start Date`,
         "EndDate" = `End Date`) %>%
  mutate(StartDate = as.character(StartDate),
         EndDate = as.character(EndDate),
         StartMonth = strsplit(StartDate, split = "-")[[1]][2],
         EndMonth = strsplit(EndDate, split = "-")[[1]][2],
         StartDay = strsplit(StartDate, split = "-")[[1]][3],
         EndDay = strsplit(EndDate, split = "-")[[1]][3]) %>%

# Get rid of original StartDate and EndDate columns  
  select(-StartDate, -EndDate)
```

## Combine athletes and region
```{r}
athletes2 <- athletes %>%
  left_join(regions, by = c('NOC' = 'NOC')) %>%
  select(-notes, -NOC, -Games)
```


## Combine olymweat2 and athletes and filter out indoor sports
```{r}
# Join athletes2 and olymweat2 dataset
olympics1 <- left_join(athletes2, olymweat2, by = c("Year" = "Year", "Season" = "Season")) %>%
  select(-City.y) %>%
  rename(City = City.x) %>%

# Filter out Olympics that do not follow modern format
  filter(!(Year %in% c(1896,1900,1904,1906,1908,1912,1920,1924,1928) & Season == "Summer")) %>%
  mutate(CityTemp1 = TempMon1,
         CityTemp2 = TempMon2) %>%

# Join inout dataset that contains information about indoor/outdoor sports    
  left_join(inout, by = c("Sport" = "Sport")) %>%

# Filter out indoor sports
  filter(isindoor == 0) %>%
  
# Select out columns without further use
  select(-ID, -Sex, -Age, -Height, -Weight, -Team, -Event, -TempMon1, -TempMon2)
```


## Combine olympics1 with countrytemp
```{r}
olympics2 <- left_join(olympics1, countrytemp, by = c("region" = "name"))

# create vector of the starting and ending months of each olympic
startmon <- as.character(as.integer(olympics1$StartMonth))
endmon <- as.character(as.integer(olympics1$EndMonth))

# Loop through olympics2 to gather the temperature of months that corresponds to the olympic months for each athlete's home country
# Store this data in histtemp1 and histtemp2
histtemp1 <- c()
histtemp2 <- c()
for (i in 1:nrow(olympics2)){
  histtemp1[i] <- olympics2[[i,startmon[i]]]
  histtemp2[i] <- olympics2[[i,endmon[i]]]
}

# Create new column corresponding to historic temperature of home countries in each Olympics
olympics2$histtemp1 <- histtemp1
olympics2$histtemp2 <- histtemp2

# Select out columns without further use
olympics2 <- olympics2 %>%
  select(-`1`,-`2`,-`3`,-`4`,-`5`,-`6`,-`7`,-`8`,-`9`,-`10`,-`11`,-`12`, -Annual_temp)
```


## Calculate difference between host city temperature and home country temperature
```{r}
# Calculate the ratio of days in the first and second month
olympics3 <- olympics2 %>%
  mutate(NumDayMon1 = ifelse(as.integer(EndDay) > `Number of Days`, 
                             `Number of Days`, 
                             `Number of Days` - as.integer(EndDay)),
         NumDayMon2 = `Number of Days` - NumDayMon1,
         RatioMon1 = NumDayMon1/`Number of Days`,
         RatioMon2 = 1 - RatioMon1,
         CityTemp2 = ifelse(is.na(CityTemp2), 0, CityTemp2)) %>%

# Calculate the total difference
  mutate(tempdiff = (CityTemp1 - histtemp1)*RatioMon1 + (CityTemp2 - histtemp2)*RatioMon2) %>%
  
# Select out columns without further use 
  select(-CityTemp1, -CityTemp2, -histtemp1, -histtemp2, -NumDayMon1, 
         -NumDayMon2, -RatioMon1, -RatioMon2, -`Number of Days`, -StartMonth, 
         -EndMonth, -StartDay, -EndDay, -isindoor)
```

## Combine elevation data with olympics3 and compute elevation difference
```{r}
# Join olympics3 and elevation dataset
olympics4 <- olympics3 %>%
  left_join(elevation, by = c("region" = "Country")) %>%

# Calculate average elevation difference
  mutate(elevdiff = Altitude - Elevation) %>% 

# Select out columns without further use
  select(-Elevation, -Altitude)
```

## Clean up medal data
```{r}
# Quantify each medals as follows: Gold = 3, Silver = 2, Bronze = 1, None = 0
olympics5 <- olympics4 %>%
  mutate(Medal = ifelse(is.na(Medal), "0", Medal),
         Medal = ifelse(Medal == "Bronze", "1", Medal),
         Medal = ifelse(Medal == "Silver", "2", Medal),
         Medal = ifelse(Medal == "Gold", "3", Medal),
         Medal = as.integer(Medal)) 
```  

## Create dataset that contains total medals for each olympics
```{r}
# Group by Year and Season and take the sum of Medal
totalmedals <- olympics5 %>%
  group_by(Year,Season) %>%
  summarise(TotMed = sum(Medal))
```

## Calculate percentage of medals won by each country
```{r}
# Calculate the sum of all medals won by each country in each olympics
olympics6 <- olympics5 %>%
  group_by(Year, Season, region) %>%
  summarise(medalswon = sum(Medal)) %>%
  
# Calculate the proportion of total medals that each country won in each olympics 
  left_join(totalmedals, by = c("Year" = "Year", "Season" = "Season")) %>%
  mutate(medratio = medalswon / TotMed)
```

## Get rid of distinctions between each athletes and focus on country level performance
```{r}
# Get rid of all individual distinctions between athletes and focus on country
olympics7 <- olympics5 %>% 
  select(Year, Season, region, tempdiff, elevdiff)

# There are multiple athletes from the same countries in every olympics, creating duplicates
# Find which rows has duplicates in olympics7
duplicated <- olympics7 %>%
  duplicated() 

# Get rid of duplicates
olympics8 <- olympics7[!duplicated,]
```

## Combined join medalratio data onto temperature/elevation data
```{r}
# Join dataset that contains medalratio info onto olympics8
olympics9 <- olympics6 %>% 
  left_join(olympics8, by = c("Year" = "Year", "Season" = "Season", "region" = "region")) %>%
  
# Select out columns that does not have further use  
  select(-medalswon, -TotMed)
```


## save olympics9 as R.data
```{r}
save(olympics9, "olympics9", file = "olympics9.Rdata")
```

















